\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2022-12-15 sections/introduction.tex
Some small fixes

Dynamically-typed programming languages such as JavaScript and Python have risen in popularity in recent decades \cite{chatley_next_2019}. These languages present a low barrier of entry, especially to beginner programmers: they require no type declaration, variable types or object structures can be modified dynamically, and functions can deal with dynamic input using ad-hoc polymorphism and runtime reflection. However, studies show that dynamically-typed languages negatively affect development productivity \cite{kleinschmager_static_2012}, code usability \cite{mayer_empirical_2012}, and code quality \cite{gao_type_2017, ray_large-scale_2017, meyerovich_empirical_2013}. They are often found to produce error-prone code \cite{chen_empirical_2020, wang_empirical_2015,xu_python_2016} and require strong programmer discipline to avoid pitfalls \cite{chen_empirical_2020}. For these reasons, many modern dynamically-typed languages have introduced static typing annotations as part of the core language features in recent years (e.g.\ \textit{TypeScript}~\cite{microsoft_javascript_nodate} and \textit{mypy}~\cite{mypy_mypy_nodate}).

Functional programming languages have long enjoyed rigorous type systems and expressive type-level features. Techniques such as type inference and algebraic types have been standard practice for decades in functional languages such as ML and Haskell, and more recently in multi-paradigm languages, such as Rust and TypeScript. Haskell is considered a ``type-system laboratory" \cite{hudak_history_2007}. Various type system advances were introduced in Haskell and ended up in mainstream languages years or even decades after. Type classes, an implementation of generic programming, were introduced to Haskell in 1988~\cite{hudak_history_2007}, and now can be found in most popular languages such as C\#~\cite{bill_wagner_constraints_2022}, Java~\cite{oracle_generic_2022}, and TypeScript~\cite{microsoft_documentation_2022}.

The challenge of programming in statically-typed languages is that type errors can sometimes be difficult to resolve~\cite{tirronen_understanding_2015, hage_solved_2020}. In particular, they may point to locations that are not the real error causes, expose errors in cryptic language, or provide misleading fixing suggestions~\cite{wu_how_2017}.
% Compared to dynamic type systems, static type systems offer programmers opportunities to weed out a large number of errors at compile time (reducing the need for run-time debugging) and to increase usability~\cite{mayer_static_2012} and codebase maintainability~\cite{kleinschmager_static_2012}. In recent years, the programming community has seen a trend of migrating away from dynamically typed codebases due to the lack of maintainability and refactoring safety when systems reach large scale and complexity~\cite{chatley_next_2019}. These migration methods include the introduction of gradual type annotations (e.g., JavaScript to TypeScript or Python with mypy) or transition to a modern strongly-typed language (Scala, Rust, Haskell, PureScript, etc.). However, with dynamically typed languages (e.g., JavaScript, PHP, and Python) still being most peoples' formative experience with programming and with such languages now firmly entrenched in education~\cite{stackoverflow_stack_2022}, this transition may pose difficulties for programmers who lack experience writing type-safe programs.

@@

This paper introduces \chameleon{}, an interactive type debugging tool for Haskell. It can visualize the relevant context of a type error (where it happens or could have happened and which parts of code cause it). In addition, \chameleon{} allows programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve ambiguity. The most noticeable features provided by \chameleon{} are the type compare tool, the deduction step feature, and the candidate expression card feature. These features are integrated into a unified debugging environment and can be enabled or disabled separately based on the programmers' preferences and debugging needs. \chameleon{} is open-source and is available in both web~\cite{anonymous_chameleon_2022} and desktop versions.  
% \todo{Introduce Haskell earlier?  Motivate WHY Haskell?} The current implementation of \chameleon{} targets the Haskell 2010 language standard. However, it is planned to extend it to support other programming languages using the same underlying ideas and techniques. Historically, Haskell has been a test bed for advanced language features, and it is common for features established in Haskell to be transferred to mainstream programming languages.

>>>>

Dynamically typed programming languages such as JavaScript and Python have risen in popularity in recent decades \cite{chatley_next_2019}. These languages present a low barrier of entry, especially to beginner programmers: they require no type declaration, variable types or object structures can be modified dynamically, and functions can deal with dynamic input using ad-hoc polymorphism and runtime reflection. However, studies show that dynamically typed languages negatively affect development productivity \cite{kleinschmager_static_2012}, code usability \cite{mayer_empirical_2012}, and code quality \cite{gao_type_2017, ray_large-scale_2017, meyerovich_empirical_2013}. They are often found to produce error-prone code \cite{chen_empirical_2020, wang_empirical_2015,xu_python_2016} and require strong programmer discipline to avoid pitfalls \cite{chen_empirical_2020}. For these reasons, many modern dynamically-typed languages have introduced static typing annotations as part of the core language features in recent years (e.g.\ \textit{TypeScript}~\cite{microsoft_javascript_nodate} and \textit{mypy}~\cite{mypy_mypy_nodate}).

Functional programming languages have long enjoyed rigorous type systems and expressive type-level features. Techniques such as type inference and algebraic types have been standard practice for decades in functional languages such as ML and Haskell, and more recently in multi-paradigm languages, such as Rust and TypeScript. Various type system advances were introduced in Haskell and ended up in mainstream languages years or even decades after, leading many to consider Haskell the ``type-system laboratory" \cite{hudak_history_2007}.  Type classes, an implementation of generic programming, were introduced to Haskell in 1988~\cite{hudak_history_2007}, and now can be found in most popular languages such as C\#~\cite{bill_wagner_constraints_2022}, Java~\cite{oracle_generic_2022}, and TypeScript~\cite{microsoft_documentation_2022}.

One crucial challenge of programming in statically-typed languages is that type errors can sometimes be difficult to resolve~\cite{tirronen_understanding_2015, hage_solved_2020}. In particular, they may point to locations that are not the root  causes of the type error, expose errors in cryptic language, or provide misleading fixing suggestions~\cite{wu_how_2017}.
% Compared to dynamic type systems, static type systems offer programmers opportunities to weed out a large number of errors at compile time (reducing the need for run-time debugging) and to increase usability~\cite{mayer_static_2012} and codebase maintainability~\cite{kleinschmager_static_2012}. In recent years, the programming community has seen a trend of migrating away from dynamically typed codebases due to the lack of maintainability and refactoring safety when systems reach large scale and complexity~\cite{chatley_next_2019}. These migration methods include the introduction of gradual type annotations (e.g., JavaScript to TypeScript or Python with mypy) or transition to a modern strongly-typed language (Scala, Rust, Haskell, PureScript, etc.). However, with dynamically typed languages (e.g., JavaScript, PHP, and Python) still being most peoples' formative experience with programming and with such languages now firmly entrenched in education~\cite{stackoverflow_stack_2022}, this transition may pose difficulties for programmers who lack experience writing type-safe programs.

@@

This paper introduces \chameleon{}, an interactive type debugging tool for Haskell. It can visualize the relevant context of a type error: where it happens or could have happened and which parts of the code cause it. In addition, \chameleon{} allows programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve ambiguity. The most noticeable features are the type compare tool (Section \ref{sub:type-compare}), the candidate expression card (Section \ref{sub:candidate-expression}), and the deduction step (Section \ref{sub:deduction-steps}). These features are integrated into a debugging environment and can be enabled or disabled separately based on the programmers' preferences and debugging needs. \chameleon{} is open-source and is available in web~\cite{anonymous_chameleon_2022} and desktop versions.  
% \todo{Introduce Haskell earlier?  Motivate WHY Haskell?} The current implementation of \chameleon{} targets the Haskell 2010 language standard. However, it is planned to extend it to support other programming languages using the same underlying ideas and techniques. Historically, Haskell has been a test bed for advanced language features, and it is common for features established in Haskell to be transferred to mainstream programming languages.

### 2022-12-15 sections/related-work.tex
Updated bib
Some small fixes

\subsection{Finding all type error locations}
Many have studied the approach of finding all locations that contribute to a type error~\cite{stuckey_interactive_2003, haack_type_2004, pavlinovic_practical_2015, schilling_constraint-free_2012}. Type error slicing~\cite{haack_type_2004} is a technique that finds locations that are complete and minimal for the type error. Internally labeled constraints and Minimal Unsatisfiable Set (MUS) manipulation are used to generate these slices. The language supported in Haack's work was a subset of Standard ML. The original Chameleon~\cite{stuckey_interactive_2003} used a similar technique but generate constraints in Constraint handling rules (CHR). As a result,  Chameleon supported advanced type-level features (type classes and functionally dependent types). The project also introduced the ability to query type information through a command line-style interface. Although Chameleon was firmly grounded in results from type theory, its designs were never evaluated with user studies. While finding all error locations is useful in comprehending type errors, it is only 1 of the 7 properties listed in the proposed manifesto of good type error reporting~\cite{yang_improved_2000}.

>>>>

\subsection{Finding all type error locations}
Many have studied the approach of finding all locations that contribute to a type error~\cite{stuckey_interactive_2003, haack_type_2004, pavlinovic_practical_2015, schilling_constraint-free_2012}. Type error slicing~\cite{haack_type_2004} is a technique that finds locations that are complete and minimal for the type error. Internally labeled constraints and Minimal Unsatisfiable Set (MUS) manipulation are used to generate these slices. The language supported in Haack's work was a subset of Standard ML. The original Chameleon~\cite{stuckey_interactive_2003} used  Constraint handling rules (CHR) to support the computing of type error slices in Haskell. Chameleon also supported advanced type-level features (type classes and functionally dependent types). The project also introduced the ability to query type information through a command line-style interface. Although Chameleon was firmly grounded in results from type theory, its designs were never evaluated with user studies. While finding all error locations is useful in comprehending type errors, it is only 1 of the 7 properties listed in the proposed manifesto of good type error reporting~\cite{yang_improved_2000}.
To the best of our knowledge, ours is the first user-centered evaluation of an interactive type debugging system involving type-error slicing.

### 2022-12-15 sections/walkthrough.tex
Some small fixes

\subsection{Basic mode} \label{sub:basic}
Maxine writes a function to calculate the sum of a list of
numbers, but \chameleon{} shows there is a type error (Fig.~\ref{fig:basic-mode-1}). 
After reading the error reports, Maxine realizes that the error possibly revolves 
around the expression \texttt{xs}.

@@

After scanning the type error report, Maxine realizes that \texttt{xs} can be
either \texttt{[a]} or \texttt{Int}. By matching the color in the
conflicting type panel (Fig. \ref{fig:anatomy}-G) and the highlighted error locations 
Maxine knows that the \texttt{[a]} results from the pattern matching of the
\texttt{:} operator, while \texttt{Int} results from using \texttt{+} to
 add two expressions. 
At this point, Maxine knows the possible type 1 aligns with her understanding of
the program, and therefore, the error locations with blue highlights must be
erroneous. After examining the program, it is obvious that Maxine missed
applying the \texttt{sum} function recursively at the right-hand side of the
addition.

@@

Maxine writes additional code to add only even numbers in a list of integers, reusing the \texttt{sum} function she wrote
earlier. After saving the file, \chameleon{} shows that there is a type error in the
expression \texttt{sum} (Fig.~\ref{fig:balance-mode-1}). However, this is not   \todo{Add annotation to figure like prev ones i.e. (1) so can refer to in text without reader searching for it?}
helpful because Maxine knows that the implementation of  \texttt{sum} is correct 
because it was verified in the previous task. Switching to
balanced mode, Maxine noticed that \chameleon{} shows two cards: \texttt{sum}
and \texttt{evens}.

@@

\begin{figure}
   \centering
        \includegraphics[width=\linewidth,trim=0mm 8mm 0mm 0mm]{images/balanced-mode-2.pdf}
        \caption{
            Clicking on the \texttt{evens} card (5) results in the changes in the
            conflicting types panel to show the possible types for \texttt{evens},
            and the changes highlights color to reflect the assumption that the
            definition of \texttt{evens} is the cause of the error.  
        }
        \label{fig:balance-mode-2}
\end{figure}

Maxine therefore clicks on the \texttt{evens} card and \chameleon{} shows two
possible types for the expression \texttt{[Int]} and \texttt{[Int] -> [Int]}
(Fig. \ref{fig:balance-mode-2}). Knowing the expression \texttt{evens} holds
a temporary list of even integers (hence it is of \texttt{[Int]} types), Maxine
knows the Possible type 2 is unintended, and blue in-text highlights must
contain the cause. It does not take long until Maxine realizes that the list 
\texttt{l} is not supplied to the \texttt{filter} function.

@@

For the task shown in section \ref{sub:balanced}, if Maxine is not satisfied by
the options provided by \chameleon{}, by switching to advanced
mode, she has access to the debugging steps and the explanation layer.

>>>>

\subsection{Basic mode} \label{sub:basic}
Maxine writes a function to calculate the sum of a list of
numbers, but \chameleon{} shows there is a type error (Fig.~\ref{fig:basic-mode-1}). 
After reading the error reports, Maxine realizes that the error revolves 
around the expression \texttt{xs}. That is: \texttt{xs} can be
either \texttt{[a]} or \texttt{Int}. By matching the color in the
conflicting type block (Fig. \ref{fig:anatomy}-H) and the highlighted error locations 
Maxine knows that the \texttt{[a]} results from the pattern matching of the
\texttt{:} operator, while \texttt{Int} results from using \texttt{+} to
 add two expressions.

@@

At this point, Maxine knows the possible type 1 aligns with her intention, and therefore, the error locations with blue highlights must be erroneous. After examining the program, it comes clear that Maxine forgets to apply the \texttt{sum} function recursively at the right-hand side of the addition.

@@

Maxine writes additional code to add only even numbers in a list of integers, reusing the \texttt{sum} function she wrote earlier. After saving the file, \chameleon{} shows a type error in the expression \texttt{sum} (Fig.~\ref{fig:balance-mode-1}). However, this is not helpful because Maxine has just verified the implementation of \texttt{sum}. Switching to balanced mode, \chameleon{} shows two cards: \texttt{sum} and \texttt{evens}.

@@

\begin{figure}
   \centering
        \includegraphics[width=\linewidth,trim=0mm 8mm 0mm 0mm]{images/balanced-mode-2.pdf}
        \caption{
            Clicking on the \texttt{evens} card (5) results in the changes in the
            conflicting types panel to show the possible types for \texttt{evens},
            and the changes highlight color to reflect the assumption that the
            definition of \texttt{evens} is the cause of the error.  
        }
        \label{fig:balance-mode-2}
\end{figure}

Maxine therefore clicks on the \texttt{evens} card and \chameleon{} reports two
possible types for the expression \texttt{[Int]} and \texttt{[Int] -> [Int]}
(Fig. \ref{fig:balance-mode-2}). Knowing the expression \texttt{evens} holds
a temporary list of even integers (hence it is of \texttt{[Int]} types), Maxine
knows the Possible type 2 is unintended. The locations with blue highlights must
contain the cause. It does not take long for Maxine to realize the list 
\texttt{l} is not supplied to the \texttt{filter} function.

@@

For the task shown in section \ref{sub:balanced}, if Maxine is not satisfied by
the options provided by \chameleon{}, by switching to advanced
mode, she has access to the debugging steps.
